The Second Data Release of the INT Photometric H$\alpha$ Survey of the Northern Galactic Plane (IPHAS DR2)

### 2014-04-16 paper/iphas-dr2.tex

\begin{abstract} % BACKGROUND, OBJECTIVE, METHODS, RESULTS, CONCLUSIONS
The INT/WFC Photometric H$\alpha$ Survey 
of the Northern Galactic Plane (IPHAS)
is a 1800~deg$^2$ imaging survey
covering Galactic latitudes $|b| < 5\degr$
and longitudes $\ell=30\degr$ to 215\degr\ 
in the $r$, $i$ and \ha\ filters 
using the Wide Field Camera (WFC) 
on the 2.5-metre Isaac Newton Telescope (INT) in La Palma.
We present the first quality-controlled and
globally calibrated source catalogue
derived from the survey,
providing single-epoch photometry
for 219 million unique sources
across 92 per cent of the footprint.
The observations were carried out between 2003 and 2013
at a median seeing of 1.1~arcsec (sampled at 0.33~arcsec/pixel)
and to a mean $5\sigma$-depth of 
21.2~($r$), 20.0~($i$) and 20.3~($\ha$)
in the Vega magnitude system.
We explain the data reduction 
and quality control procedures,
describe and test the global re-calibration,
and detail the construction of the new catalogue.
We show that the new calibration is accurate to
0.03~mag (rms) 
and recommend a series of quality criteria
to select the most reliable data
from the catalogue.
Finally, we demonstrate the ability of the 
catalogue's unique
$(r-\ha,\ r-i)$ diagram to
(i) characterise stellar populations and extinction regimes
towards different Galactic sight-lines
and (ii) select \ha\ emission-line objects.
IPHAS is the first survey to offer comprehensive CCD photometry
of point sources across the Galactic Plane at visible wavelengths,
providing the much-needed counterpart
to recent infrared surveys.
\end{abstract}

@@

\section{Introduction}
The INT/WFC Photometric H$\alpha$ Survey
of the Northern Galactic Plane \citep[IPHAS;][]{Drew2005}
is providing new insights into the contents and structure of the disk of the Milky Way.
This large-scale programme of observation
-- spanning almost a decade 
and using more than 300 nights in competitive open time 
at the Isaac Newton Telescope (INT) in La Palma --
aims to provide the digital update 
to the photographic northern H$\alpha$ surveys 
of the mid-20th century \citep[see][]{Kohoutek1999}. 
By increasing the sensitivity 
with respect to these previous surveys 
by a factor $\sim$1000 (7~magnitudes), 
IPHAS can expand
the limited bright samples of Galactic emission line objects 
previously available into larger, deeper, and more statistically-robust samples that will 
better inform our understanding 
of the early and late stages of stellar evolution.
Since the publication of the Initial Data Release \citep[IDR;][]{Gonzalez-Solares2008},
these aims have begun to be realised through a
range of published studies including: 
a preliminary catalogue of candidate emission-line objects \citep{Witham2008};
discoveries of new symbiotic stars \citep{Corradi2008, Corradi2010, Corradi2011}; 
new cataclysmic variables \citep{Witham2007,Wesson2008,Aungwerojwit2012}; 
new groups of young stellar objects
\citep{Vink2008,Barentsen2011a,Wright2012};
new classical Be stars \citep{Raddi2013};
along with discoveries of new supernova remnants \citep{Sabin2013}
and new and remarkable planetary nebulae 
\citep{Mampaso2006, Viironen2009a, Viironen2009b, Sabin2010, Corradi2011, Viironen2011}.

Over the years it has become apparent that the legacy of IPHAS 
will reach beyond these traditional \ha\ applications 
of identifying emission-line stars and nebulae. 
Through the provision of $r$ and $i$ broadband photometry 
alongside narrowband \ha\ data,
IPHAS has created the opportunity 
to study Galactic Plane populations 
in a new way.
For example, the survey's unique $(r-\ha,\ r-i)$ colour-colour
diagram has been shown to provide simultaneous constraints 
on intrinsic stellar colour and interstellar extinction \citep{Drew2008}. 
This has opened the door 
to a wide range of Galactic science applications, 
including the mapping of extinction across the Plane in three dimensions
and the probabilistic inference of stellar properties
\citep{Sale2009, Sale2010, Giammanco2011, Sale2012, Barentsen2013}. 
In effect, the availability of narrowband \ha\
alongside $r$ and $i$ magnitudes
provides coarse spectral information for huge samples of stars 
which are otherwise too faint or numerous 
to be targeted by spectroscopic surveys.
For such science applications to succeed, however, 
it is vital that the imaging data are transformed 
into a homogeneously-calibrated photometric catalogue, 
in which quality problems 
and duplicate detections are flagged.

@@

\begin{figure*}
        \includegraphics[width=1\linewidth]{figures/footprint/footprint_large.jpg}
        \caption{Survey area showing the footprints
        of all the quality-approved IPHAS fields
        which have been included in this data release.
        The area covered by each field has been coloured black
        with a semi-transparent opacity of 20 per cent,
        such that regions where fields overlap are darker.
        The IPHAS strategy is to observe each field twice
        with a small offset,
        and hence the vast majority of the area 
        is covered twice (dominant grey colour).
        There are small overlaps between all the neighbouring fields
        which can be seen as a honeycomb pattern
        of dark grey lines across the survey area.
        Regions with incomplete data are apparent as white gaps (no data) 
        or in light grey (denoting that one offset is missing).
        The dark vertical strip near $\ell \simeq$125\degr\ 
        is an arbitrary consequence of the tiling pattern,
        which was populated starting from 0h in right ascension.}
        \label{fig:footprint}
\end{figure*}

@@

The WFC is a mosaic of 4 CCDs 
that captures a sky area of close to 0.29~deg$^2$.
To cover the Northern Plane with some overlap,
the survey area was divided into 7,635 telescope pointings.
Each of these pointings is accompanied by an offset position
displaced by $+$5 arcmin in Dec 
and $+$5 arcmin in RA,
to deal with inter-CCD gaps, detector imperfections,
and to enable quality checks. 
Hence, the basic unit of observation
amounts to $2 \times 3$ exposures, 
in which each of the 3 survey filters is exposed at 2 offset sky positions 
within, typically, an elapsed time of 10 minutes.
We shall refer to the unit of 3 exposures at the same position 
as a \emph{field},
and the combination of two fields at a small offset as a \emph{field pair}.
Altogether, the survey contains 15,270 fields
grouped into 7,635 field pairs.
To achieve the desired survey depth
of 20th magnitude or fainter in each filter, 
the exposure times were set at 120~s (\ha), 
30~s ($r$) and 10~s ($i$)
in the vast majority of the survey observations.\footnote{The $r$-band exposure time was 10~s instead of 30~s in the first months of data taking. Since October 2010, the $i$-band exposure time 
has been increased from 10~s to 20~s to by-pass a sporadic exposure timing bug that affects the WFC.}

@@

Fig.~\ref{fig:footprint} shows the footprint
of the quality-approved observations included in this work. 
The fields which remain missing 
-- covering 8 per cent of the survey area --
are predominantly located towards the Galactic anti-centre 
at $\ell > 120\degr$.
Fields at these longitudes are mainly accessed from La Palma 
in the months of November-December,
which is when the La Palma weather and seeing conditions are often poor,
forcing many (unsuccessful) repeat observations.
To enable the survey to be brought to completion, 
a decision was made recently to limit repeats in this area 
to individual fields requiring replacement,
i.e. fresh observations in all 3 filters may only be obtained 
at one of the two offset positions.  
The catalogue is structured such that it is clear 
where a contemporaneous observation of both halves of a field pair
is available.

@@

All raw IPHAS data were transferred
to the Cambridge Astronomical Survey Unit (CASU) 
for initial processing and archival.
The procedures used by CASU were originally devised
for the INT Wide Field imaging Survey \citep[WFS;][]{McMahon2001,Irwin2005},
which was a 200 deg$^2$ survey programme carried out 
between 1998 and 2003 after the WFC was commissioned.
Because IPHAS uses the same telescope and camera combination,
we have been able to benefit from the existing WFS pipeline.
A detailed description of the processing steps 
is found in \citet{Irwin2001}.
Its application to IPHAS has previously been described
by \citet{Drew2005} and \citet{Gonzalez-Solares2008},
and the source code is available
on-line\footnote{http://casu.ast.cam.ac.uk/surveys-projects/software-release}. 
In brief, the pipeline takes care of bias subtraction,
linearity correction, flat-fielding,
gain correction and de-fringing.

@@

At the time of preparing DR2,
the CASU pipeline had processed
74,195 single-band IPHAS exposures 
in which a total of 1.9~billion \emph{candidate} detections
were made at the default detection level of 1.25\,$\sigma$.
This total inevitably includes spurious objects, artefacts and
duplicate detections;
in \S\ref{sec:catalogue} we will explain
how these have been removed or flagged in the final catalogue.
The pipelined data set -- comprising 2.5~terabyte of FITS files --
was then transferred to the University of Hertfordshire
for the purpose of transforming the raw
detection tables into a more reliable source catalogue which is 
(i) quality-controlled,
(ii) homogeneously calibrated, and 
(iii) contains user-friendly columns and warning flags.
It is these post-processing steps which are explained next.

@@

(1) \emph{Depth.} 
We discarded any exposures for which the $5\sigma$ limiting magnitude 
was brighter than 20th magnitude in the $r$-band
or brighter than 19th in $i$ or \ha. 
Such data were typically obtained during poor weather or full moon.
Most observations were significantly better than these limits.
Fig.~\ref{fig:depth} presents the distribution of limiting magnitudes
for all quality-approved fields;
the mean depths and standard deviations are 
$21.2\pm0.5$ ($r$), $20.0\pm0.3$ ($i$) and $20.3\pm0.3$ (\ha).
The depth achieved depended
most strongly on the presence of the moon,
which was above the horizon during 62 per cent 
of the observations.
The great range in sky brightness this
produced is the main reason for the wide and bi-modal shape
of the $r$-band limiting magnitude distribution 
(top panel in Fig.~\ref{fig:depth}).
In contrast, the depths attained in $i$ and \ha\ 
are less sensitive to moonlight, leading to
narrower magnitude limit distributions
(middle and bottom panels in Fig.~\ref{fig:depth}).
To a lesser extent, the wide spread in the $r$-band depth
is also explained by the shorter exposure time
that was used for this band during the first months of data-taking.

@@

(3) \emph{Seeing.} 
The original survey goal was to obtain data 
at a resolution better than 1.7~arcsec,
as evaluated by measuring the average PSF Full Width at Half Maximum (FWHM)
across the detectors,
which is a proxy for the image quality
arising from the combined effects of atmospheric and dome seeing.
This target is currently attained across 86 per cent of the footprint.
To increase the sky area offered by the data release slightly,
we have decided to accept data obtained with FWHM up to 2.5~arcsec.
Fig.~\ref{fig:seeing} presents the distribution
of the PSF FWHM for the approved fields.
In the $r$-band, 90 per cent is better than 1.5~arcsec,
50 per cent is better than 1.1~arcsec,
and 10 per cent is better than 0.8~arcsec
(the WFC pixel scale is 0.33~arcsec/px).
In \S\ref{sec:catalogue} we will explain
that the photometry compiled in the source catalogue
is normally derived from the field with the
best-available seeing for a given object,
and that the FWHM measurement
is available as a column in the catalogue.

@@

(4) \emph{Photometric repeatability.} 
The IPHAS field-pair observing strategy normally
ensures that every pointing is immediately followed 
by an offset pointing at a displacement
of $+$5~arcmin in Dec and $+$5~arcmin in RA.
This allows pairs of images to be checked 
for the presence of clouds or electronic noise.
To exploit this information,
the overlap regions of all field pairs were systematically cross-matched
to verify the consistency of the photometry
for stars observed in both pointings.
We automatically rejected field pairs
in which more than 2 per cent of the stars 
showed an inconsistent measurement at the level of 0.2~mag,
or more than 25 per cent were inconsistent at the level of 0.1~mag.
These limits were set empirically after inspecting
the images and photometry by eye.

@@

(7) \emph{Contemporaneous field data.} 
Finally, only exposures which are part of a sequence 
of three consecutive images of the same field
(\ha, $r$, $i$)
were considered for inclusion in the release. 
This ensures that the three bands for a given field
are observed contemporaneously --  
normally within 5 minutes of each other.
%An exception was made for 9 fields where the three exposures 
%could not be obtained within the same night
%but for which the time gap between exposures did not exceed 48 hours.
We note that the source catalogue details the exact epoch
at the start of each exposure
(columns \emph{rMJD}, \emph{iMJD}, \emph{haMJD}).

@@

For the purpose of providing an initial calibration 
of the $r$ and $i$ broadband fluxes,
photometric standard fields were observed every night.
The standards were chosen from a list based on 
the \cite{Landolt1992} and Stetson (http://cadcwww.dao.nrc.ca/standards) 
objects.
Two or three standard fields were observed 
during the evening and morning twilight,
and at intervals of 2-3 hours throughout the night.
The CASU pipeline automatically identified the observed standards 
and used them to determine a sigma-clipped average zeropoint \textsc{magzpt}
for each night and filter,
such that the number counts $DN$ 
in the pipeline-corrected CCD frames
relate to a magnitude $m$ as:
\begin{equation}
\begin{split}
   m  = & \textsc{magzpt} - 2.5 \log_{10}( DN / \textsc{exptime} ) \\
 &  - \textsc{extinct}\cdot(\textsc{airmass}-1) - \textsc{apcor} - \textsc{percorr},
\label{eqn:mag}
\end{split}
\end{equation}
where \textsc{exptime} is the exposure time in seconds,
\textsc{extinct} is the atmospheric extinction coefficient 
(set in the pipeline at 0.09 for $r$ and 0.05 for $i$
as representative averages for La Palma),
\textsc{airmass} is the normalised optical path length 
through the atmosphere,
and \textsc{apcor} is a correction for the flux
lost outside of the aperture used.
Finally, \textsc{percorr} is a correction based on the difference
between the median dark sky for a CCD against the median for all the CCDs, 
and as such is an ancillary correction 
to account for sporadic gain variations;
it is zero for most exposures.
All these quantities correspond to header keywords in the 
multi-extension FITS files produced by the CASU pipeline.

The broadband zeropoints were determined such that the resulting magnitude system
refers to the Spectral Energy Distribution (SED) of Vega 
as the zero colour object. 
Colour equations were used to transform between the IPHAS passbands 
and the Johnson-Cousins system 
of the published standard star photometry.
The entire procedure has been found to deliver zeropoints which 
are accurate at the level of 1--2 per cent 
in stable photometric conditions \citep{Gonzalez-Solares2011}.

Unlike the broadbands, 
standard-star photometry is not available in the literature 
for the \ha\ passband
and hence there is no formally recognised flux scale 
for it.
We can specify here, however, 
that the detected flux for Vega in the
IPHAS \ha\ filter is 3.14 magnitudes
less than the flux captured by the much broader
$r$ band which includes the \ha\ band.
Hence to assure zero colour relative to the broadbands,
we set the default zeropoint for the narrowband to be:
\begin{equation}
\textsc{magzpt}_{H\alpha} = \textsc{magzpt}_r - 3.14.
\label{eqn:zpha}
\end{equation}
For reference, Table~\ref{tbl:flux} details the flux of Vega
in the IPHAS filter system.
We refer to \citet{Drew2005} for a more comprehensive
discussion on the throughput curves of the bands.

\begin{table}
    \caption{Mean monochromatic flux of Vega in the IPHAS filter system,
             defined as 
             $\langle f_\lambda\rangle~=~\int{f_\lambda(\lambda) S(\lambda) \lambda d \lambda} / \int{S(\lambda)\lambda d \lambda}$,
             where $S(\lambda)$ is the photon response function 
             (which includes atmospheric transmission,
             filter transmission
             and CCD response)
             and $f_\lambda(\lambda)$ is the model SED for Vega
             by Kurucz (http://kurucz.harvard.edu/stars.html).
             For reference, we also provide the 
             filter equivalent width EW~=~$\int S(\lambda) d\lambda$,
             the mean photon wavelength 
             $\lambda_0~=~\int{S(\lambda)\lambda d\lambda} / \int{S(\lambda)d\lambda}$,
             and the pivot wavelength
             $\lambda_p~=~\sqrt{\int{S(\lambda)\lambda d\lambda} / \int{\frac{S(\lambda)}{\lambda} d\lambda}}$.
             These notations follow the definitions by \citet{Bessell2012}.
             When multiplying $\langle f_\lambda\rangle$ by EW,
             we find that the flux received for Vega
             is 3.14~mag less in \ha\ than in $r$. 
          }
    \label{tbl:flux}
    \begin{center}
    \begin{tabular}{lccrc}
    \toprule
     Filter  & $\langle f_\lambda \rangle$  & \multicolumn{1}{c}{EW} &$\lambda_0$ & $\lambda_p$    \\
             & [erg\,cm$^{-2}$\,s$^{-1}$\,\AA$^{-1}$] & [\AA] & [\AA] & [\AA] \\
    \midrule
$r$ &  $2.47 \times 10^{-9}$  &  785.6 &  6223  &   6211   \\
\ha &  $1.81 \times 10^{-9}$ &   59.6 &  6568  &   6568    \\
$i$  &  $1.30 \times 10^{-9}$  &  759.9  &  7674  &   7674   \\
    \bottomrule
    \end{tabular}
    \end{center}
\end{table}

@@

Despite the best efforts made to obtain a nightly calibration,
large surveys naturally possess field-to-field variations
due to atmospheric changes during the night
and imperfections in the pipeline or the instrument
(e.g. the WFC is known to suffer from sporadic errors
in the timing of exposures).
This is demonstrated in Fig.~\ref{fig:ccd_before},
where we show the combined colour-colour diagram
for all 2,801 quality-approved fields
which are located towards the Galactic anti-centre
(defined here as $160\degr < \ell < 200\degr$).
The main locus of stars is poorly defined
in the diagram due to the presence of
incorrectly calibrated fields.
The calibration errors can exceed 0.1~mag or more
and need to be corrected for 
during a global re-calibration procedure.
For comparison, Fig.~\ref{fig:ccd_after}
shows the identical diagram which was constructed
after our re-calibration procedure.

>>>>

\begin{abstract} % BACKGROUND, OBJECTIVE, METHODS, RESULTS, CONCLUSIONS
The INT/WFC Photometric H$\alpha$ Survey 
of the Northern Galactic Plane (IPHAS)
is a 1800~deg$^2$ imaging survey
covering Galactic latitudes $|b| < 5\degr$
and longitudes $\ell=30\degr$ to 215\degr\ 
in the $r$, $i$ and \ha\ filters 
using the Wide Field Camera (WFC) 
on the 2.5-metre Isaac Newton Telescope (INT) in La Palma.
We present the first quality-controlled and
globally-calibrated source catalogue
derived from the survey,
providing single-epoch photometry
for 219 million unique sources
across 92 per cent of the footprint.
The observations were carried out between 2003 and 2012
at a median seeing of 1.1~arcsec (sampled at 0.33~arcsec/pixel)
and to a mean $5\sigma$-depth of 
21.2~($r$), 20.0~($i$) and 20.3~($\ha$)
in the Vega magnitude system.
We explain the data reduction 
and quality control procedures,
describe and test the global re-calibration,
and detail the construction of the new catalogue.
We show that the new calibration is accurate to
0.03~mag (rms) 
and recommend a series of quality criteria
to select the most reliable data
from the catalogue.
Finally, we demonstrate the ability of the 
catalogue's unique
$(r-\ha,\ r-i)$ diagram to
(i) characterise stellar populations and extinction regimes
towards different Galactic sight-lines
and (ii) select \ha\ emission-line objects.
IPHAS is the first survey to offer comprehensive CCD photometry
of point sources across the Galactic Plane at visible wavelengths,
providing the much-needed counterpart
to recent infrared surveys.
\end{abstract}

@@

\section{Introduction}
The INT/WFC Photometric H$\alpha$ Survey
of the Northern Galactic Plane \citep[IPHAS;][]{Drew2005}
is providing new insights into the contents and structure of the disk of the Milky Way.
This large-scale programme of observation
-- spanning a decade so far 
and using more than 300 nights in competitive open time 
at the Isaac Newton Telescope (INT) in La Palma --
aims to provide the digital update 
to the photographic northern H$\alpha$ surveys 
of the mid-20th century \citep[see][]{Kohoutek1999}. 
By increasing the sensitivity 
with respect to these preceding surveys 
by a factor $\sim$1000 (7~magnitudes), 
IPHAS can expand
the limited bright samples of Galactic emission line objects 
previously available into larger, deeper, and more statistically-robust samples that will 
better inform our understanding 
of the early and late stages of stellar evolution.
Since the publication of the Initial Data Release \citep[IDR;][]{Gonzalez-Solares2008},
these aims have begun to be realised through a
range of published studies including: 
a preliminary catalogue of candidate emission-line objects \citep{Witham2008};
discoveries of new symbiotic stars \citep{Corradi2008, Corradi2010, Corradi2011}; 
new cataclysmic variables \citep{Witham2007,Wesson2008,Aungwerojwit2012}; 
new groups of young stellar objects
\citep{Vink2008,Barentsen2011a,Wright2012};
new classical Be stars \citep{Raddi2013};
along with discoveries of new supernova remnants \citep{Sabin2013}
and new and remarkable planetary nebulae 
\citep{Mampaso2006, Viironen2009a, Viironen2009b, Sabin2010, Corradi2011, Viironen2011}.

Over the years it has become apparent that the legacy of IPHAS 
will reach beyond these traditional \ha\ applications 
of identifying emission-line stars and nebulae. 
Through the provision of $r$ and $i$ broadband photometry 
alongside narrowband \ha\ data,
IPHAS has created the opportunity 
to study Galactic Plane populations 
in a new way.
For example, the survey's unique $(r-\ha,\ r-i)$ colour-colour
diagram has been shown to provide simultaneous constraints 
on intrinsic stellar colour and interstellar extinction \citep{Drew2008}. 
This has opened the door 
to a wide range of Galactic science applications, 
including the mapping of extinction across the Plane in three dimensions
and the probabilistic inference of stellar properties
\citep{Sale2009, Sale2010, Giammanco2011, Sale2012, Barentsen2013}. 
In effect, the availability of narrowband \ha\
alongside $r$ and $i$ magnitudes
provides coarse spectral information for huge samples of stars 
which are otherwise too faint or numerous 
to be targeted by spectroscopic surveys (cf. the use of
Stromgren $uvbyH\beta$ photometry at blue wavelengths).
For such science applications to succeed, however, 
it is vital that the imaging data are transformed 
into a homogeneously-calibrated photometric catalogue, 
in which quality problems 
and duplicate detections are flagged.

@@

\begin{figure*}
        \includegraphics[width=1\linewidth]{figures/footprint/footprint_large.jpg}
        \caption{Survey area showing the footprints
        of all the quality-approved IPHAS fields
        which have been included in this data release.
        The area covered by each field has been coloured black
        with a semi-transparent opacity of 20 per cent,
        such that regions where fields overlap are darker.
        The IPHAS strategy is to observe each field twice
        with a small offset,
        and hence the vast majority of the area 
        is covered twice (dominant grey colour).
        There are small overlaps between all the neighbouring fields
        which can be seen as a honeycomb pattern
        of dark grey lines across the survey area.
        Regions with incomplete data are apparent as white gaps (no data) 
        or in light grey (indicating that one offset is missing).
        The dark vertical strip near $\ell \simeq$125\degr\ 
        is an arbitrary consequence of the tiling pattern,
        which was populated starting from 0h in Right Ascension.}
        \label{fig:footprint}
\end{figure*}

@@

The WFC is a mosaic of 4 CCDs 
that captures a sky area of close to 0.29~deg$^2$ at a pixel scale of
0.33 arcsec/pixel.
To cover the Northern Plane with some overlap,
the survey area was divided into 7,635 telescope pointings.
Each of these pointings is accompanied by an offset position
displaced by $+$5 arcmin in Dec 
and $+$5 arcmin in RA,
to deal with inter-CCD gaps, detector imperfections,
and to enable quality checks. 
Hence, the basic unit of observation
amounts to $2 \times 3$ exposures, 
in which each of the 3 survey filters is exposed at 2 offset sky positions 
within, typically, an elapsed time of 10 minutes.
We shall refer to the unit of 3 exposures at the same position 
as a \emph{field},
and the combination of two fields at a small offset as a \emph{field pair}.
Altogether, the survey contains 15,270 fields
grouped into 7,635 field pairs.
To achieve the desired survey depth
of 20th magnitude or fainter in each filter, 
the exposure times were set at 120~s (\ha), 
30~s ($r$) and 10~s ($i$)
in the vast majority of the survey observations.\footnote{The $r$-band exposure time was 10~s instead of 30~s in the first months of data taking. Since October 2010, the $i$-band exposure time 
has been increased from 10~s to 20~s to by-pass a sporadic exposure timing bug that affects the WFC.}

@@

Fig.~\ref{fig:footprint} shows the footprint
of the quality-approved observations included in this work. 
The fields which remain missing 
-- covering 8 per cent of the survey area --
are predominantly located towards the Galactic anti-centre 
at $\ell > 120\degr$.
Fields at these longitudes are mainly accessed from La Palma 
in the months of November-December,
which is when the La Palma weather and seeing conditions are often poor,
forcing many (unsuccessful) repeat observations.
To enable the survey to be brought to completion, 
a decision was made recently to limit repeats in this area 
to individual fields requiring replacement,
i.e. fresh observations in all 3 filters may only be obtained 
at one of the two offset positions.  
The catalogue is structured such that it is clear 
where contemporaneous observations of both halves of a field pair
are available.

@@

All raw IPHAS data were transferred
to the Cambridge Astronomical Survey Unit (CASU) 
for initial processing and archival.
The procedures used by CASU were originally devised
for the INT Wide Field imaging Survey \citep[WFS;][]{McMahon2001,Irwin2005},
which was a 200 deg$^2$ survey programme carried out 
between 1998 and 2003.
Because IPHAS uses the same telescope and camera combination,
we have been able to benefit from the existing WFS pipeline.
A detailed description of the processing steps 
is found in \citet{Irwin2001}.
Its application to IPHAS has previously been described
by \citet{Drew2005} and \citet{Gonzalez-Solares2008},
and the source code is available
on-line\footnote{http://casu.ast.cam.ac.uk/surveys-projects/software-release}. 
In brief, the pipeline takes care of bias subtraction,
linearity correction, flat-fielding,
gain correction and de-fringing.

@@

At the time of preparing DR2,
the CASU pipeline had processed
74,195 single-band IPHAS exposures 
in which a total of 1.9~billion \emph{candidate} detections
were made at the default detection level of 1.25\,$\sigma$.
This total inevitably includes spurious objects, artefacts and
duplicate detections;
in \S\ref{sec:catalogue} we will explain
how these have been removed or flagged in the final catalogue.
The pipelined data set -- comprising 2.5~terabyte of FITS files --
was then transferred to the University of Hertfordshire
for the purpose of transforming the raw
detection tables into a more reliable source catalogue which
(i) is quality-controlled,
(ii) is homogeneously calibrated, and 
(iii) contains user-friendly columns and warning flags.
It is these post-processing steps which are explained next.

@@

(1) \emph{Depth.} 
We discarded any exposures for which the $5\sigma$ limiting magnitude 
was brighter than 20th magnitude in the $r$-band
or brighter than 19th in $i$ or \ha. 
Such data were typically obtained during poor weather or full moon.
Most observations were significantly better than these limits.
Fig.~\ref{fig:depth} presents the distribution of limiting magnitudes
for all quality-approved fields;
the mean depths and standard deviations are 
$21.2\pm0.5$ ($r$), $20.0\pm0.3$ ($i$) and $20.3\pm0.3$ (\ha).
The depth achieved depended
most strongly on the presence of the moon,
which was above the horizon during 62 per cent 
of the observations.
The great range in sky brightness this
produced is behind the wide and bi-modal shape
of the $r$-band limiting magnitude distribution 
(top panel in Fig.~\ref{fig:depth}).
In contrast, the depths attained in $i$ and \ha\ 
are less sensitive to moonlight, leading to
narrower magnitude limit distributions
(middle and bottom panels in Fig.~\ref{fig:depth}).
To a lesser extent, the wide spread in the $r$-band depth
is also explained by the shorter exposure time
that was used for this band during the first months of data-taking.

@@

(3) \emph{Seeing.} 
The original survey goal was to obtain data 
at a resolution better than 1.7~arcsec,
as evaluated by measuring the average PSF Full Width at Half Maximum (FWHM)
across the detectors.
% Stating the obvious ...a new insertion? which is a proxy for the image quality
% arising from the combined effects of atmospheric and dome seeing.
This target is currently attained across 86 per cent of the footprint.
To increase the sky area offered by the data release slightly,
we have decided to accept data obtained with FWHM up to 2.5~arcsec.
Fig.~\ref{fig:seeing} presents the distribution
of the PSF FWHM for the approved fields.
In the $r$-band, 90 per cent is better than 1.5~arcsec,
50 per cent is better than 1.1~arcsec,
and 10 per cent is better than 0.8~arcsec.
% slipped in earlier (the WFC pixel scale is 0.33~arcsec/px).
In \S\ref{sec:catalogue} we will explain
that the photometry compiled in the source catalogue
is normally derived from the field with the
best-available seeing for a given object,
and that the FWHM measurement
is available as a column in the catalogue.

@@

(4) \emph{Photometric repeatability.} 
The IPHAS field-pair observing strategy normally
ensures that every pointing is immediately followed 
by an offset pointing at a displacement
of $+$5~arcmin in Dec and $+$5~arcmin in RA.
This allows pairs of images to be checked 
for the presence of clouds or electronic noise.
To exploit this information,
the overlap regions of all field pairs were systematically cross-matched
to verify the consistency of the photometry
for stars observed in both pointings.
We automatically rejected field pairs
in which more than 2 per cent of the stars 
showed an inconsistent measurement at the level of 0.2~mag,
or more than 26 per cent were inconsistent at the level of 0.1~mag.
These limits were set empirically after inspecting
the images and photometry by eye.

@@

(7) \emph{Contemporaneous field data.} 
Finally, only exposures which are part of a sequence 
of three consecutive images of the same field
(\ha, $r$, $i$)
were considered for inclusion in the release. 
This ensures that the three bands for a given field
are observed contemporaneously --  
nearly always within 5 minutes of each other.
%An exception was made for 9 fields where the three exposures 
%could not be obtained within the same night
%but for which the time gap between exposures did not exceed 48 hours.
We note that the source catalogue details the exact epoch
at the start of each exposure
(columns \emph{rMJD}, \emph{iMJD}, \emph{haMJD}).

@@

For the purpose of providing an initial calibration 
of the $r$ and $i$ broadband fluxes,
photometric standard fields were observed every night.
The standards were chosen from a list based on 
the \cite{Landolt1992} and Stetson (http://cadcwww.dao.nrc.ca/standards) 
objects.
Two or three standard fields were observed 
during the evening and morning twilight,
and at intervals of 2-3 hours throughout the night.
The CASU pipeline automatically identified the observed standards 
and used them to determine a sigma-clipped average zeropoint \textsc{magzpt}
for each night and filter,
such that the number counts $DN$ 
in the pipeline-corrected CCD frames
relate to a magnitude $m$ as:
\begin{equation}
\begin{split}
   m  = & \textsc{magzpt} - 2.5 \log_{10}( DN / \textsc{exptime} ) \\
 &  - \textsc{extinct}\cdot(\textsc{airmass}-1) - \textsc{apcor} - \textsc{percorr},
\label{eqn:mag}
\end{split}
\end{equation}
where \textsc{exptime} is the exposure time in seconds,
\textsc{extinct} is the atmospheric extinction coefficient 
(set in the pipeline at 0.09 for $r$ and 0.05 for $i$ as representative
averages for the telescope site),
\textsc{airmass} is the normalised optical path length 
through the atmosphere and
\textsc{apcor} is a correction for the flux
lost outside of the aperture used
(we adopt a 2\farcs 3-diameter circular aperture by default).
Finally, \textsc{percorr} is a correction based on the difference
between the median dark sky for a CCD against the median for all the CCDs, 
and as such is an ancillary correction 
to account for sporadic gain variations --
it is zero for most exposures.
All these quantities correspond to header keywords in the 
multi-extension FITS files produced by the CASU pipeline.

The broadband zeropoints were determined such that the resulting magnitude system
refers to the spectral energy distribution (SED) of Vega 
as the zero colour object. 
Colour equations were used to transform between the IPHAS passbands 
and the Johnson-Cousins system 
of the published standard star photometry.
The entire procedure has been found to deliver zeropoints which 
are accurate at the level of 1--2 per cent 
in stable photometric conditions \citep{Gonzalez-Solares2011}.

Unlike the broadbands, 
standard-star photometry is not available in the literature 
for the \ha\ passband
and hence there is no formally recognised flux scale 
for it.
We can specify here, however, 
that the detected flux for Vega in the
IPHAS \ha\ filter is 3.14 magnitudes
less than the flux captured by the much broader
$r$ band which includes the \ha\ band.
Hence to assure zero colour relative to the broadbands,
we set the default zeropoint for the narrowband to be:
\begin{equation}
\textsc{magzpt}_{H\alpha} = \textsc{magzpt}_r - 3.14.
\label{eqn:zpha}
\end{equation}
For reference, Table~\ref{tbl:flux} details the flux of Vega
in the IPHAS filter system.
Data on the throughput curves of the bands can be obtained from
the Isaac Newton Group website\footnote{http://catserver.ing.iac.es/filter/list.php?instrument=WFC where the
filters are named WFCH6568, WFCSloanR and WFCSloanI}.

\begin{table}
    \caption{Mean monochromatic flux of Vega in the IPHAS filter system,
             defined as 
             $\langle f_\lambda\rangle~=~\int{f_\lambda(\lambda) S(\lambda) \lambda d \lambda} / \int{S(\lambda)\lambda d \lambda}$,
             where $S(\lambda)$ is the photon response function 
             (which includes atmospheric transmission,
             filter transmission
             and CCD response)
             and $f_\lambda(\lambda)$ is the CALSPEC SED for Vega
             \citep{Bohlin2014}.
             For reference, we also provide the 
             filter equivalent width EW~=~$\int S(\lambda) d\lambda$,
             the mean photon wavelength 
             $\lambda_0~=~\int{S(\lambda)\lambda d\lambda} / \int{S(\lambda)d\lambda}$,
             and the pivot wavelength
             $\lambda_p~=~\sqrt{\int{S(\lambda)\lambda d\lambda} / \int{\frac{S(\lambda)}{\lambda} d\lambda}}$.
             These notations follow the definitions by \citet{Bessell2012}.
             After multiplying $\langle f_\lambda\rangle$ by the EW,
             we find that the detected flux for Vega in \ha\ is 3.14~magnitudes less
             than that received in $r$, 
          }
    \label{tbl:flux}
    \begin{center}
    \begin{tabular}{lccrc}
    \toprule
     Filter  & $\langle f_\lambda \rangle$  & \multicolumn{1}{c}{EW} &$\lambda_0$ & $\lambda_p$    \\
             & [erg\,cm$^{-2}$\,s$^{-1}$\,\AA$^{-1}$] & [\AA] & [\AA] & [\AA] \\
    \midrule
$r$ &  $2.47 \times 10^{-9}$  &  785.6 &  6223  &   6211   \\
\ha &  $1.81 \times 10^{-9}$ &   59.6 &  6568  &   6568    \\
$i$  &  $1.30 \times 10^{-9}$  &  759.9  &  7674  &   7661   \\
    \bottomrule
    \end{tabular}
    \end{center}
\end{table}

@@

Despite the best efforts made to obtain a nightly calibration,
large surveys naturally possess field-to-field variations
due to atmospheric changes during the night
and imperfections in the pipeline or the instrument
(e.g. the WFC is known to suffer from sporadic errors
in the timing of exposures).
This is demonstrated in Fig.~\ref{fig:ccd_before},
where we show the combined colour-colour diagram
for the set of 2\,801 quality-approved fields
which are located towards the Galactic anti-centre
($160\degr < \ell < 200\degr$).
Despite the fact that all the fields included in the diagram
have been quality-approved,
the main locus of stars is poorly defined
due to the presence of calibration errors,
which regularly exceed 0.1~mag.
For comparison, Fig.~\ref{fig:ccd_after}
shows the identical diagram
obtained after re-calibrating.
